---
title: "Week 3 Analysis - Logistic Regression"
author: "Juho Lähteenmaa"
date: "13 November 2018"
---

# Analysis, IODS week 3

## Import data

Read data from own data folder.

```{r}
alc <- read.csv("C:/Users/juhol/OneDrive/Documents/GitHub/IODS-project/data/alc.csv", header = T, sep = ",")

alc <- alc[,-1] # First column only duplicates the row numbers

dim(alc)

str(alc)

summary(alc)
```

Data icludes 382 observations and 35 variables. Data is combined from two datasets which are loaded from [here][id1]. Site includes also the meta text for data which is usefull to read. R-code for data wrangling can be found in [here][id2] and data is available in [here][id3]. 
 
[id1]:https://archive.ics.uci.edu/ml/machine-learning-databases/00320/
[id2]:https://github.com/juholaht/IODS-project/blob/master/data/create_alc.R
[id3]:https://github.com/juholaht/IODS-project/blob/master/data/alc.csv


## Analysis

Packages used in this section:

```{r, error=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(GGally)
library(cowplot)
library(boot)

```


Lets introduce the column names of our data.

```{r}
colnames(alc)
```

In our analyse section we are interested in relationships between alcohol consumption and social parameters. Lets choose variable "high use of alcohol" (binary) as the dependent variable of interest and "extra-curricular activities" (binary), "romantic relationship" (binary), "quality of family relationships" (numeric: from 1 - very bad to 5 - excellent) and "going out with friends" (numeric: from 1 - very low to 5 - very high) as social parameters of interest. Lets also control the age and sex.

```{r}
# Lets make subset of data with variables "sex", "age", "activities", "romantic", "famrel", "goout" and "high_use".

variables_sub <- c("sex", "age", "activities", "romantic", "famrel", "goout", "high_use")

alc_sub <- select(alc, one_of(variables_sub))

str(alc_sub)

```

Lets plot alcohol use which each of these variables.

```{r}
alc_sex <- ggplot(data = alc_sub, aes(x = sex, fill = high_use)) + geom_bar() + ggtitle("Consumption of alcohol and sex")

alc_sex


alc_age <- ggplot(data = alc_sub, aes(x = age, fill = high_use)) + geom_bar() + ggtitle("Consumption of alcohol and age")

alc_age

alc_act <- ggplot(data = alc_sub, aes(x = activities, fill = high_use)) + geom_bar() + ggtitle("Consumption of alcohol and activities")

alc_act

alc_rom <- ggplot(data = alc_sub, aes(x = romantic, fill = high_use)) + geom_bar() + ggtitle("Consumption of alcohol and romantic relationship")

alc_rom


alc_family <- ggplot(data = alc_sub, aes(x = famrel, fill = high_use)) + geom_bar() + ggtitle("Consumption of alcohol and quality of family relationship")

alc_family


alc_goout <- ggplot(data = alc_sub, aes(x = goout, fill = high_use)) + geom_bar() + ggtitle("Consumption of alcohol and going out with friends")

alc_goout




```

It looks clear that high consuming of alcohol is more common for males than for females.
Interesting fact in this sample is that all age groups from 15 to 18 (where we have better sample) are relatively having same part of high users. It also looks clear that in the group of students having extra-curricular activities, the heavy using is not as common as in other group. Only by looking the histogram, it looks that better quality of family relationships decreases the risk of high consumption of alcohol. Not suprisingly the group of students who are spending more time in out with friends is also consuming more alcohol.

Lets looking this in more detail. Lets start from GLM-model
$$\text{HighUse}_{i} = \beta_{0} + \beta_{1}\text{Sex}_{i} + \beta_{2}\text{Age}_{i} + \beta_{3}\text{Activities}_{i} + \beta_{4}\text{Romantic}_{i} + \beta_{5}\text{Family}_{i} + \beta_{6}\text{GoOut}_{i} + \epsilon_{i}$$
and then using the backward method for finding the best candidate for the best linear model.
```{r}
# First model
glm1 <- glm(high_use ~ sex + age + activities + romantic + famrel + goout, data = alc_sub, family = "binomial")

step(glm1, direction = "backward")
```

By Akaike information criterion the best model is the following:
$$\text{HighUse}_{i} = \beta_{0} + \beta_{1}\text{Sex}_{i} + \beta_{2}\text{Activities}_{i} + \beta_{3}\text{Family}_{i} + \beta_{4}\text{GoOut}_{i} + \epsilon_{i}$$

This model has $\text{AIC} = 400.6$.

```{r}
glm_best <- glm(formula = high_use ~ sex + activities + famrel + goout, family = "binomial", 
    data = alc_sub)

summary(glm_best)
```
We can calculate $\text{McFadden's Pseudo} R^2$ representing the overall effect size of the model (more about from [here][id4]):

[id4]: http://www.glmj.org/archives/articles/Smith_v39n2.pdf

```{r}
ll.null <- glm_best$null.deviance/-2
ll.proposed <- glm_best$deviance/-2

pseudo_r_sqr <- (ll.null-ll.proposed)/ll.null

pseudo_r_sqr



ll.null2 <- glm1$null.deviance/-2
ll.proposed2 <- glm1$deviance/-2

pseudo_r_sqr_org <- (ll.null2-ll.proposed2)/ll.null2

pseudo_r_sqr_org



```

We get that $\text{McFadden's Pseudo} R^2 = 0.16$ which is relatively small. The original model has $\text{McFadden's Pseudo} R^2 = 0.17$ so having extra variables does not increase the effect size significantly. 


In this model all other variables (including the intercept) but "extra-curricular activities" (binary) are statistically significant (counted by Wald-test) in 99 % confidence level. Also variable "extra-curricular activities" is significant in 90 % confidence level. All the coefficients are in "log(odds)" -format. With binary variables, coefficient is represinting the "log(odds ratio)" -format, for example variable sex has $log(\text{odd ratio}) = 0.9971 \Leftrightarrow \text{odd ratio} = 2.71041$. The social variables which were the variables of our interest, having extra-curricular activities has an negative effect to high consumption of alcohol as the good family relationships also does, both with log(odd) -estimation around -0.40 - -0.50. Going out with friends has a positive effect to heavy drinking, $log(\text{odd}) = 0.8118 \Leftrightarrow \text{odd} = 2.251958$, meaning that approximately the ratio between heavy drinkers of those who go out and those how don't is $\frac{5}{2}$.

We can provide 2x2 cross tabulation of predictions versus the actual values. Lets also have a visual look up, how our model "fitted" with data.
```{r}
## Predicted values of our model
predicted_data <- data.frame(
  probability_of_hc = glm_best$fitted.values,
  high_consumption = alc_sub$high_use)
 
predicted_data <- mutate(predicted_data, prediction = probability_of_hc > 0.5)

predicted_data <- predicted_data[  order(predicted_data$probability_of_hc, decreasing=FALSE),]

predicted_data$rank <- 1:nrow(predicted_data)

# head(predicted_data)

# tabulate the target variable versus the predictions
table(high_use = predicted_data$high_consumption, prediction = predicted_data$prediction)

 
# Plotting
ggplot(data=predicted_data, aes(x=rank, y=probability_of_hc)) +
  geom_point(aes(color=high_consumption), alpha=1, shape=4, stroke=2) +
  xlab("Index") +
  ylab("Predicted probability of high consumption of alcohol")
 



```


We see that our model with social parameters controlled by sex made average quality predictions, data is mainly distributed in right way. It can be assumed from the small overall effect size of the model (only 0.17 as shown above) that it would not perfectly generate the right predictions. Also what we have to take into account is that the model doesn't tell us about the causalitys. For example the causality between high consumption of alcohol and going out with friends has probably a causality in both ways.

Lets do this also with cross-validation for the model.

```{r}
# define a loss function (average prediction error)
loss_func <- function(class, prob) {
  n_wrong <- abs(class - prob) > 0.5
  mean(n_wrong)
}

# Average number of wrong predictions in the (training) data

loss_func(predicted_data$high_consumption, predicted_data$probability_of_hc)

# 10-fold cross-validation

cv <- cv.glm(data = alc_sub, cost = loss_func, glmfit = glm_best, K = 10)

# Average number of wrong predictions in the cross validation
cv$delta[1]

# For comparing, lets try probabilities P(High_use) = 1 and 0 

# P(High_use) = 1
loss_func(predicted_data$high_consumption, prob = 1)

# P(High_use) = 0
loss_func(predicted_data$high_consumption, prob = 0)



```

Average number of wrong predictions in the training data is 0.23 and 10-fold cross-validation gives 0.24. Our model gives better predictions than just assuming the probability of high consumption to be zero (loss function gives 0.3) or one (loss function gives 0.7). 