---
title: "Week 4 Analysis - Clustering and classification"
author: "Juho Lähteenmaa"
date: "20 November 2018"

---

# Analysis, IODS week 4


Here are the packages used in this session:

```{r, error=FALSE, warning=FALSE, message=FALSE}
# REMOVE ALL OBJECTS
rm(list=ls())

# access the MASS package
library(MASS)
library(corrplot)
library(tidyverse)
library(ggplot2)
library(GGally)
library(plotly)
```


## Data


Let's load the Boston data from the MASS package. Data is all about the "Housing Values in Suburbs of Boston". More about the data can be read from [here][id1].

[id1]: https://stat.ethz.ch/R-manual/R-devel/library/MASS/html/Boston.html

```{r}


# load the data
data("Boston")

# explore the dataset
str(Boston)

summary(Boston)

# plot matrix of the variables
ggpairs(Boston, lower = "blank", upper = list(continuous = "points", combo =
  "facethist", discrete = "facetbar", na = "na"))

```


Boston dataset has 506 observations and 14 variables. All variables are in numeric format. Variable *chas* is a dummy variable (Charles River dummy variable, 1 if tract bounds river and 0 otherwise) and variable *rad* is an index (index of accessibility to radial highways).

Only by waching the plot above, it looks that the data has some linear relations. These relations can bee seen with correlation matrix.

```{r}
# calculate the correlation matrix and round it
cor_matrix<-cor(Boston) 

# print the correlation matrix
cor_matrix %>% round(digits = 2)

# visualize the correlation matrix
##corrplot(cor_matrix, method="circle", type = "upper", cl.pos = "b", tl.pos = "d", tl.cex = 0.6)
corrplot(cor_matrix, method="circle", tl.pos = "d")
```

Some high correlations between variables can be seen, for example between *index of accessibility to radial highways* and *full-value property-tax rate per \$10,000* (0.91), *proportion of owner-occupied units built prior to 1940* and *weighted mean of distances to five Boston employment centres* (-0.75) and *nitrogen oxides concentration* and *proportion of non-retail business acres per town* (-0.77).


## LDA

For further analysis, we will standardize the data. Let's have a same kind of lookup in the data after we have standadized the data.

```{r}

# center and standardize variables
boston_std <- scale(Boston)

# summaries of the scaled variables
summary(boston_std)

# class of the boston_scaled object
class(boston_std)

# change the object to data frame
boston_scaled <- as.data.frame(boston_std)


```

From summary we can see that we have succesfully standardized the data. All the variables have mean zero and deviation is standardized. Let's analyse the *crime rate* variable deeper.

```{r}

summary(boston_scaled$crim)


# create a quantile vector of crim and print it
bins <- quantile(boston_scaled$crim)
bins

# create a categorical variable 'crime'
crime <- cut(boston_scaled$crim, breaks = bins, include.lowest = TRUE, label = c("low", "med_low", "med_high", "high"))

# look at the table of the new factor crime
table(crime)

# remove original crim from the dataset
boston_scaled <- dplyr::select(boston_scaled, -crim)

# add the new categorical value to scaled data
boston_scaled <- data.frame(boston_scaled, crime)


```

As wee see from *crime* table above, all quantiles have same number of samples as it should be. Let's divide the dataset to train and test sets, so that 80% of the data belongs to the train set. 

```{r}
# number of rows in the Boston dataset 
n <- nrow(boston_scaled)

# choose randomly 80% of the rows
ind <- sample(n,  size = n * 0.8)

# create train set
train <- boston_scaled[ind,]

# create test set 
test <- boston_scaled[-ind,]

# save the correct classes from test data
correct_classes <- test$crime

# remove the crime variable from test data
test <- dplyr::select(test, -crime)


```

Next we will fit the linear discriminant analysis on the train set. We use the categorical crime rate as the target variable and all the other variables in the dataset as predictor variables.
```{r}


# linear discriminant analysis
lda.fit <- lda(crime ~ ., data = train)

# print the lda.fit object
lda.fit

# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "red", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col="orange", length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col= "orange", pos=3)
}

# target classes as numeric
classes <- as.numeric(train$crime)

# plot the lda results
plot(lda.fit, dimen = 2, col = classes, pch = classes)
lda.arrows(lda.fit, myscale = 1)



```


From the plot and summary above we can see that variable *index of accessibility to radial highways* has the greatest seperation weight related to other variables, both in *LD1* and *LD2* dimensions. In *LDI* coefficient for *index of accessibility to radial highways* is *3.4* when second greatest weight, *0.37*, is variables *nitrogen oxides concentration*. In dimension *LD2* variable *index of accessibility to radial highways* does not have such a dominance as in *LD1*. In both dimensions it is positively correlated with crime rate. In *LD2* coefficient for it is *0.9*, when for example variables *proportion of residential land zoned for lots over 25,000 sq.ft.* and *nitrogen oxides concentration (parts per 10 million)* have both coefficient around *0.7* as in absolute value, but these *LD2* components have oppposite directions (*nitrogen oxides concentration (parts per 10 million)* increases criminal rate). In our LD1-LD2 map moving south-east inreases the criminal rate.

For next we will test how well our model predicts the *crime rate*. Earlier in our analysis we separated our data in learning and testing parts and we removed categorial variable *crime* from test data.

```{r}
# predict classes with test data
lda.pred <- predict(lda.fit, newdata = test)

# cross tabulate the results
table(correct = correct_classes, predicted = lda.pred$class)




```

In table above we see that our model predicts high crime rate very well. Also in other rate classes our model makes quite good predictions, even in classes med_high and med_low.


## K-means clustering

Let's use original standardized data with original *crime rate* variable. Then we will alalyse data with K-means clustering method. 

```{r}
boston_std <- as.data.frame(boston_std)


boston_dist <- dist(boston_std)

set.seed(123)

# determine the number of clusters
k_max <- 10

# calculate the total within sum of squares
twcss <- sapply(1:k_max, function(k){kmeans(boston_std, k)$tot.withinss})

# visualize the results
qplot(x = 1:k_max, y = twcss, geom = 'line')




```


Based on graph above we decide to use two clusters.
```{r}

# K-means clustering
km <-kmeans(boston_std, centers = 2)



# plot the Boston dataset with clusters
#pairs(Boston, col = km$cluster)
class(km$cluster)
ggpairs(boston_std, mapping = aes(col = as.factor(km$cluster), alpha = 0.3), lower = "blank", upper = list(continuous = "points", combo =
  "facethist", discrete = "facetbar", na = "na"))

```

From the plot above we see that k-means clustering works quite well, in most cases those two clusters are enough to seperate the data to two resonable distributions. For example, in the case of *nitrogen oxides concentration (nox)* we see quite clearly two different distributions.

## Bonus

For a bonus analyse we will search for a reasonable k-mean clustering for original, standardized Manhattan data. Then we will use these clusters as target variables in new LDA-model and find the variables which have most weight in clustering.

```{r}
boston_scaled_j <- scale(Boston)

boston_scaled_j <- as.data.frame(boston_scaled_j)

# calculate the total within sum of squares
twcss_j <- sapply(1:k_max, function(k){kmeans(boston_scaled_j, k)$tot.withinss})

# visualize the results
qplot(x = 1:k_max, y = twcss_j, geom = 'line')

## Lets use 7 clusters

km_j <-kmeans(boston_scaled_j, centers = 7)

lda.fit_j <- lda(km_j$cluster ~ ., data = boston_scaled_j)

lda.fit_j


# the function for lda biplot arrows
lda.arrows_j <- function(x, myscale = 1, arrow_heads = 0.1, color = "red", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric
classes_j <- as.numeric(km_j$cluster)

# plot the lda results
plot(lda.fit_j, dimen = 2, col = classes_j, pch = classes_j)
lda.arrows(lda.fit_j, myscale = 1)
```

Data is well clustered in seven part. Once again variable *rad*, *index of accessibility to radial highways* has a major influence in LD1 dimension. In LD2 based dimension variable *tax, full-value property-tax rate per \$10,000* has the greatest weight. 


## Super Bonus

```{r}

model_predictors <- dplyr::select(train, -crime)

# check the dimensions

dim(model_predictors)
dim(lda.fit$scaling)


# matrix multiplication

matrix_product <- as.matrix(model_predictors) %*% lda.fit$scaling
matrix_product <- as.data.frame(matrix_product)


plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type= 'scatter3d', mode='markers', color = train$crime)


# Let's find k-clusters for train set

n_k <- nrow(Boston)

ind_k <- sample(n_k, size = n_k*0.8)


train_k <- Boston[ind_k,]

train_k <- scale(train_k)


twcss2 <- sapply(1:k_max, function(k){kmeans(stats::na.omit(train_k), k)$tot.withinss})

# visualize the results
qplot(x = 1:k_max, y = twcss2, geom = 'line') # 2

# k-means clustering
km_k <-kmeans(train_k, centers = 2)

plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type= 'scatter3d', mode='markers', color = km_k$cluster)

# Let's increase the number of clusters to 4

km_l <-kmeans(train_k, centers = 4)

plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type= 'scatter3d', mode='markers', color = km_l$cluster)

```

In this part we can compare the 3D scatterplots where base vectors are from LDA model made above. All the points in these three 3D scatterplots are in the same cordinates, only the colours differ.The first 3D scatterplot is colored with four crime rates, the second with 2-mean clustering method and the third with 4-mean clustering method. Colours distributions in k-mean clustering plots do not respond the colour distibution of crime rate distribution. It might be that criminal rate does not have as high weight as some other variables in the sense of clustering because it has relatively slight correlation with other variables (this can be seen above).